Design and Implementation of a Multithreaded Virtual Machine for Executing Linear Logic Programs

### 2014-07-25 virtual_machine.tex
Update.

We developed a compiler that compiles LM programs to byte-code and a multi-threaded
virtual machine (VM) using the Pthreads library to run the byte-code.
The goal of our system is to keep the threads as busy as possible and to reduce inter-thread communication.

The load balancing aspect of the system is performed by our work scheduler that is based on a work
stealing algorithm. Initially, the system will partition the application
graph of $N$ nodes into $P$ subgraphs (the number of threads) and then each thread will work on their own subgraph.
During execution, threads can steal nodes of other threads to keep themselves busy.

Reduction of inter-thread communication is achieved by first ordering the node addresses present
in the code in such a way that closer nodes are clustered together and then partitioning them
to threads. During compilation, we take note of predicates that are used in rules for
communication rules (arguments with type \emph{node}) and then build a graph of nodes from
the program's axioms. The nodes of the graph are then ordered by using a breadth-first search
algorithm that changes the nodes of addresses to the domain $[0, n[$, where $n$ is the number of
nodes. Once the VM starts, we simply partition the range
$[0, n[$.

@@

When the VM starts, it reads the byte-code file and starts all threads.
The partitioning of nodes is performed by dividing the domain of node addresses across the number of available threads.
In Fig.~\ref{fig:overview} we present the overview of a virtual machine running 2 threads and a program with 6 nodes.
Node that the dotted arrows represent the edges between nodes. In each thread space, we place the nodes owned by each thread.
Initially, the threads grab their own nodes and assign the \texttt{owner} property of each.
Because only one thread is allowed to do computation on a node at any giving time, the owner property
defines the thread with such permission.

\begin{figure*}[ht]
   \centering
   \scalebox{0.60}{\includegraphics[]{overview.pdf}}
   \caption{Overview of the virtual machine (showing 2 threads).}
   \label{fig:overview}
\end{figure*}

Included in the thread space is the \emph{Work Queue}, a single linked list containing \emph{active nodes}, i.e.,
nodes that have new facts to process. Initially, the \emph{Work Queue} is filled with all the nodes of the thread
in order to derive the axioms.

The main thread loop is shown in Fig.~\ref{code:work_loop}, where the threads
inspects their work queue for active nodes. Procedure \texttt{process\_node()} takes a node with new candidate rules
and executes them. If the work queue is empty, the thread attempts to steal one node from another thread before
becoming idle (\emph{Steal node} in Fig.~\ref{fig:overview}). Starting from a random thread, it cycles through all the threads to find one active node.
Eventually, there will be no more work to do and the threads will go idle. There is a global atomic counter, a global
boolean flag and one boolean flag for each thread that are used to detect termination.
Once a thread goes idle, it decrements the global counter and changes its flag to idle. If the counter
reaches zero, the global flag is set to idle. Since every thread will be busy-waiting and checking
the global flag, they will detect the change and stop executing.

\begin{figure}[h!]
\scriptsize\begin{Verbatim}
void work_loop(thread_id tid):
   while (true):
      current_node = NULL;
      if(has_work(tid)):
         current_node = pop_work(tid); // take node from the queue
      else:
         target_thread = random(NUM_THREADS);
         for (i = 0; i < NUM_THREADS && current_node == NULL; ++i):
            // need to steal a node
            target_thread = (target_thread + 1) % NUM_THREADS;
            current_node = steal_node_from_thread(target_thread)
      if(current_node == NULL):
         become_idle(tid);
         if(!synchronize_termination(tid)):
            return;
         become_active(tid);
      else:
         process_node(current_node, tid);
\end{Verbatim}
  \caption{Thread work loop.}
  \label{code:work_loop}
\end{figure}

@@

In Fig.~\ref{fig:overview}, we also present the internal structure of a node (node \texttt{@1}). Each node
contains the following: a database composed of \emph{Linear DB} (linear facts) and \emph{Persistent DB} (persistent facts) (further explained in \ref{sec:database});
rule matching structures (\emph{Rule Matching Structures}); and a \emph{Fact Buffer}, for storing intermediate facts coming
from other threads.

Whenever a new fact is derived through rule derivation, we need to update the data structures for the corresponding node.
This is trivial if the thread that owns the node derived the fact also. However, if that is not the case, then we have to synchronize
since multiple threads need to update the same data structures. We added a lock and a boolean flag to each node to protect the access to its data structures.
For example, in Fig.~\ref{fig:overview}, if thread 2 derives a fact to node \texttt{@1} (owned by thread 1), then thread 2 checks the node's flag
and if not activated, will lock node \texttt{@1} and then perform the required updates (\emph{New fact (1)}). If the flag is activated, it means
that thread 1 is currently executing the node, therefore it will not touch the main node data structures, but instead will add the new fact
to \emph{Fact Buffer} (\emph{New fact (2)}). The facts stored in \emph{Fact Buffer} will then be processed by thread 1 whenever the node is about to be executed.

Another thread interaction happens during fact derivation. In the example, after the fact is derived and sent to node \texttt{@1}, thread 2 also may need
to activate node \texttt{@1} by pushing it to the \emph{Work Queue} of thread 1 (\emph{Activate node}). Each node has another flag called \emph{In Queue} that indicates
if the node is currently placed in the \emph{Work Queue}. If the node is not in the queue, then we push it to the queue and then check if the target thread is currently idle.
After this synchronization point, the target thread will be active and with a new node to process.

\subsubsection{Database Data Structures}\label{sec:database}

We said before that LM rules are constrained by the first argument. Because nodes can execute
independently, our database is indexed by the node address and each sub-database does not
need to deal with synchronization issues since at any given point, only one thread will be using
the database. Note that the first argument of each fact is not stored.

The database must be implemented efficiently because during matching of rules we need
to restrict the facts using a given \emph{match object}, which fixes arguments of the target predicate to instantiated values.
Each sub-database is implemented using three kinds of data structures:

\begin{itemize}
   \item \emph{Trie Data Structures} are used exclusively to store persistent facts.
   Tries are trees where facts are indexed by the common arguments.

   \item \emph{Doubly Linked List Data Structures} are used to store linear facts.
   We use a double linked list because it is very efficient to add and remove facts.

   \item \emph{Hash Table Data Structures} are used to improve lookup when linked lists are too long and when we need to do search filtered by a fixed argument. The virtual machine decides which arguments are best to be indexed
   (see "Indexing") and then
   uses an hash table indexed by the appropriate argument. If we need to go through all the facts, we just iterate through all the facts in the table. For collisions, we use the above doubly linked list data structure.
\end{itemize}

\begin{figure}[]
   \centering
   \includegraphics[width=0.40\textwidth]{hash_table.pdf}
   \caption{\small{Hash table data structure for storing predicate \texttt{a(int,int)}.}}
   \label{fig:hash_table}
\end{figure}

Figure~\ref{fig:hash_table} shows an example for a hash table data structure
with 3 linear facts indexed by the second argument and stored as doubly linked list
in bucket \texttt{2}. Each linear fact contains the regular list pointers, a \texttt{flags} field
and the fact arguments. Those are all stored continuously to improve data
locality. One use of the \texttt{flags} field is to mark that a fact is already being used. For example,
consider the rule body \texttt{a(A,B), a(C,D) -o ...}. When we first pick a fact for \texttt{a(A, B)} from the hash table,
we mark it as being used in order to ensure that, when we retrieve facts for \texttt{a(C, D)}, the first one
cannot be used since that would violate linearity.

\subsubsection{Rule Engine}\label{rule_engine}

The rule engine decides which rules may need to be executed while taking into account rule priorities.
There are 5 main data structures for scheduling rule execution;
\texttt{Rule Queue} is the bitmap representing the rules that will be run;
\texttt{Active Bitmap} contains the rules that have enough facts to be fired;
\texttt{Inactive Bitmap} contains the rules that must be dropped from \texttt{Rule Queue};
\texttt{Predicates Bitmap} marks the newly derived facts;
and \texttt{Predicates Count} counts the number of facts per predicate.
To understand how our engine works, consider the following rules and axioms:

{\footnotesize\begin{Verbatim}
// rules
a, e(1) -o b.
a -o c.
b -o d.
e(0) -o f.
c -o e(1).

// axioms
a.
e(0).
\end{Verbatim}
}

\begin{figure*}[]
   \centering
   \scalebox{0.7}{\includegraphics[]{rule_queue1.pdf}}
   \caption{Program and rule engine data structures: before and after applying the \nth{2} rule.}
   \label{fig:rule_engine}
\end{figure*}

In our example, since we have facts \texttt{a} and \texttt{e(0)},
we will execute the second rule \texttt{a -o c}.
In order to pick such rule, we need to take the least significant bit from the \texttt{Rule Queue} bitmap.
Because the derivation is successful, we will consume \texttt{a} and derive \texttt{c}.
We thus mark the \texttt{c} predicate in the \texttt{Predicates Bitmap} and the first and second rules
in \texttt{Dropped Bitmap} since such rules are no longer applicable (\texttt{a} is gone). To update the \texttt{Rule Queue},
we remove the bits marked in \texttt{Dropped Bitmap} and add the active rules marked in \texttt{Active Bitmap} that are affected
by predicates in \texttt{Predicates Map}. The engine thus schedules the fourth and fifth rules to run (Fig.~\ref{fig:rule_engine}b).

Note that every node in the program has the same set of data structures present in Fig.~\ref{fig:rule_engine}.
We use 32 bits integers to implement bitmaps and an array of 16 bits integers to count facts, resulting in
$32 + 2P$ bytes per node, where $P$ is the number of predicates.

We do a small optimization to reduce the number of derivations of persistent facts. We
divide the program rules into two sets: \emph{persistent rules} and \emph{non persistent rules}.
Persistent rules are rules where only persistent facts are involved. We compile such rules
incrementally, that is, we attempt to fire all rules where a persistent fact is used. This is called
the \emph{pipelined semi-naive} evaluation and it originated in the P2 system~\cite{Loo-condie-garofalakis-p2}.
This evaluation method avoids excessing re-derivations of the same fact. The order of derivation does not matter for those rules, since
only persistent facts are used.

\subsection{Byte-Code}

@@

Consider a rule \texttt{!a(X,Y), b(X,Z), c(X,Y) -o d(Y)} and a database with
\texttt{!a(1,2)}, \texttt{!a(2,3)}, \texttt{b(1,3)}, \texttt{b(5,3)}, \texttt{c(1,2)}, \texttt{c(1,3)},
\texttt{c(5,3)}. Rule execution proceeds in a series of recursive loops, as follows: the first loop retrieves an
iterator for the persistent facts of \texttt{!a/2} and moves the first valid fact, \texttt{!a(1,2)},
to register 0; the inner loop retrieves linear facts that match \texttt{b(1,Z)} (from the
\emph{join constraint}) and moves \texttt{b(1,3)} to register 1; in the final
loop we move \texttt{c(1,2)} to register 2 and the body of the rule is successfully matched. Next, we
derive \texttt{d(2)}, where \texttt{2} comes from register 0.
Fig.~\ref{fig:byte_code} shows the byte-code for this example.

\begin{figure}[]
\scriptsize\begin{Verbatim}
PERSISTENT ITERATE a MATCHING TO reg 0
  LINEAR ITERATE b MATCHING TO reg 1
      (match).0=0.0
    LINEAR ITERATE c MATCHING TO reg 2
        (match).0=0.0
        (match).1=0.1
      ALLOC d TO reg 3
      MVFIELDFIELD 0.1 TO 3.0
      ADDLINEAR reg 3
      REMOVE reg 2
      REMOVE reg 1
      TRY NEXT
    NEXT
  NEXT
RETURN
\end{Verbatim}
\caption{\small{Byte-code for rule \texttt{!a(X,Y), b(X,Z), c(X,Y) -o d(Y).}}}
\label{fig:byte_code}
\end{figure}

In case of failure, we jump to the previous outer loop in order to try the next candidate fact.
If a rule matches and the head is derived, we backtrack to the inner most \emph{valid loop}, i.e.,
the first inner loop that uses linear facts or, if there are no linear facts involved, to the previous 
inner loop. We need to jump to a valid loop because we may have loops with linear facts that are now invalid.
In our example, we would jump to the
loop of \texttt{b(X,Z)} and not \texttt{c(X,Y)}, since \texttt{b(1,3)} was consumed.

\begin{figure}[]
\scriptsize\begin{Verbatim}
LINEAR ITERATE a MATCHING TO reg 0
  MVFIELDREG 0.0 TO reg 1
  MVINTREG INT 1 TO reg 2
  reg 1 INT PLUS reg 2 TO reg 3
  MVREGFIELD reg 3 TO 0.0
  UPDATE reg 0
  TRY NEXT
RETURN
\end{Verbatim}
\caption{\small{Byte-code for rule \texttt{a(N) -o a(N+1)}.}}
\label{code:update}
\end{figure}

The compiler re-orders the fact expressions used in the body in order to make execution more
efficient. For example, it forces the join constraints in rules to appear at the beginning so
that matching will fail sooner rather than later. It also does the same for constraints.
Note that for every loop, the compiler adds the \emph{match object}, which contains information
about which arguments need to match, so that runtime matching is efficient.

Our compiler also detects cases where we re-derive a linear fact with new arguments.
For example, as shown in Fig.~\ref{code:update}, the rule \texttt{a(N) -o a(N+1)}
will compile to code that reuses the old \texttt{a(N)} fact.
We use a \texttt{flags} field to mark updated nodes (presented next).

\subsection{Indexing}\label{indexing}

@@

{\scriptsize
\[
Entropy(A, F) = - \sum_{v \in values(F, A)} \frac{count(F, A = v)}{total(F)} 	\log_2 \frac{count(F, A = v)}{total(F)}
\]
}

@@

For one of the arguments to score, $Entropy(A, F)$ multiplied by the number of times it has been used for lookup must be larger than the other argument.

The argument with the best score is selected and then
a global variable called \texttt{indexing\_epoch} is updated.
In order to convert the node's linked lists into hash tables, each node also has a local variable called \texttt{indexing\_epoch}
that is compared to the global variable in order to rebuild the node database according to the new indexing
information.

@@

@@

LM supports recursive types such as lists and pairs. These complex data structures are stored in
the heap of the VM and are managed through reference counting. For instance, each list
is a \emph{cons cell} with 3 fields: \texttt{tail}, the pointer to the next element of the list;
\texttt{head}, the element stored by this element of the list; and \texttt{refs} that counts the
number of pointers to this list element in the VM. The list is deleted from the heap whenever
\texttt{refs} is decremented to 0.

>>>>

\section{The Virtual Machine}
\label{virtual_machine}

We have developed a compiler that compiles LM programs to byte-code
and a multi-threaded virtual machine (VM) using POSIX threads
that runs the byte-code.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A key goal of our design is to keep the threads as busy as possible
and to reduce inter-thread communication. When the VM starts, it reads
the byte-code file and starts all threads. Initially, the VM will
partition the application graph of $N$ nodes into $T$ subgraphs (the
number of threads) and then each thread will work on their own
subgraph. Reduction of inter-thread communication is achieved by first
ordering the node addresses present in the code in such a way that
closer nodes are clustered together and then partitioning them to
threads. During compilation, we take note of predicates that are used
in rules with communication (arguments with type \emph{node}) and
then build a graph of nodes from the program's axioms. The nodes of
the graph are then ordered by using a breadth-first search algorithm
that changes the nodes of addresses to the domain $[0, n[$, where $n$
is the number of nodes. Once the VM starts, we simply partition the
range $[0, n[$ into $T$ subgraphs.

During execution, threads can steal nodes of other threads to keep
themselves busy. The load balancing aspect of the system is performed
by our work scheduler that is based on a simple work stealing
algorithm. The pseudo-code for the main thread loop is shown in
Fig.~\ref{code:work_loop}. In each round, a thread inspects its work
queue for active nodes with new candidate rules and, if there is any,
procedure \texttt{process\_node()} executes it. If the work queue is
empty, the thread attempts to steal one node from another
thread. Starting from a random thread, it cycles through all the
threads to find one active node. Eventually, there will be no more
work to do and the threads will go idle. There is a global atomic
counter, a global boolean flag and one boolean flag for each thread
that are used to detect termination. Once a thread goes idle, it
decrements the global counter and changes its flag to idle. If the
counter reaches zero, the global flag is set to idle. Since every
thread will be busy-waiting and checking the global flag, they will
detect the change and stop executing.

\begin{figure}[ht]
{\footnotesize
\begin{Verbatim}
void thread_work_loop(thread_id tid)
   while (true)
      node = pop_node_from_work_queue(tid)
      if (node)
         process_node(tid, node)
      else
         // need to steal a node
         target = random(NUM_THREADS)
         for (i = 0; i < NUM_THREADS && !node; i++)
            target = (target + 1) % NUM_THREADS
            node = steal_node_from_thread(target)
         if (!node)
            // try to terminate
            become_idle(tid)
            if (!synchronize_termination(tid))
               return
            become_active(tid)
\end{Verbatim}
}
\caption{Thread work loop}
\label{code:work_loop}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{figures/overview.pdf}
\caption{Layout of the virtual machine}
\label{fig:overview}
\end{figure*}

Figure~\ref{fig:overview} presents the layout of our virtual machine
for a program with six nodes and two running threads. Each thread
space includes the nodes owned by the thread (the dotted arrows
represent the edges between nodes) and a \emph{Work Queue}, a single
linked list containing \emph{active nodes}, i.e., nodes that have new
facts to process. Initially, the \emph{Work Queue} is filled with all
the nodes of the thread in order to derive the
axioms. Figure~\ref{fig:overview} also illustrates the internal
structure layout of a node, which includes: the database of linear
facts (\emph{Linear DB}); the database of persistent facts
(\emph{Persistent DB}); the rule matching structures (\emph{Rule
Engine}); and a auxiliary buffer for storing intermediate facts
coming from other threads (\emph{Fact Buffer}).

Whenever a new fact is derived through rule derivation, we need to
update the data structures for the corresponding node. This is trivial
if the thread that derived the fact also owns the node. If that is not
the case, then we have to synchronize since multiple threads might be
updating the same node's data structures. We added a lock and a
boolean flag to each node to protect the access to its data
structures. When the flag is activated, it means that the node is
currently being executed by the owning thread. For example, in
Fig.~\ref{fig:overview}, if thread 2 derives a fact to
node \texttt{@1} (owned by thread 1), then thread 2 checks the node's
flag and if not activated, will lock node \texttt{@1} and perform the
required updates (\emph{New fact (1)}). If the flag is activated, it
will not touch the main node data structures, but instead will add the
new fact to \emph{Fact Buffer} (\emph{New fact (2)}). The facts stored
in \emph{Fact Buffer} will then be processed whenever the
corresponding node is about to be executed.

There is another thread interaction that might happen during fact
derivation if the node receiving a new fact is not active. In such
case, the sending thread needs to activate the node by pushing it to
the \emph{Work Queue} of the target thread. For example, consider
again the situation in which thread 2 sends a new fact to
node \texttt{@1}. If node \texttt{@1} is not active, then thread 2
also needs to activate it by pushing it to the \emph{Work Queue} of
thread 1. After this synchronization point, if the target thread is
currently idle, it will become active and with a new node to process.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Database Data Structures}
\label{sec:database}

We said before that LM rules are constrained by the first argument and
that each node has its own database of linear and persistent
facts. Moreover, since only one thread at a time will be using a
node's database, we do not need to deal with synchronization
issues. Note also that the first argument of each fact is not stored.

The database must be implemented efficiently because during matching
of rules we need to restrict the facts using a given \emph{match
object}, which fixes arguments of the target predicate to instantiated
values. Each node's database is implemented using three kinds of data
structures:

\begin{itemize}
\item \emph{Trie Data Structures} are used exclusively to store 
  persistent facts. Tries are trees where facts are indexed by common
  prefix arguments.
\item \emph{Doubly Linked List Data Structures} are used to store 
  linear facts. We use a double linked list because it is a very
   efficient way to add and remove facts.
\item \emph{Hash Table Data Structures} are used to improve lookup when 
  linked lists are too long and when we need to do search filtered by
  a fixed argument. The virtual machine decides which arguments are
  best to be indexed (see Section~\ref{indexing}) and then uses a hash table
  indexed by the appropriate argument. If we need to go through all
  the facts, we just iterate through all the facts in the table. For
  collisions, we use the above doubly linked list data structure.
\end{itemize}

Figure~\ref{fig:hash_table} shows an example for a hash table data
structure for a \texttt{a(int,int)} predicate with 3 linear facts
indexed by the second argument and stored as a doubly linked list in
bucket \texttt{2}. Each linear fact contains the regular list
pointers, a \texttt{flags} field and the fact arguments. Those are all
stored continuously to improve data locality. One use of
the \texttt{flags} field is to mark that a fact is already being
used. For example, consider the rule \texttt{a(A,B), a(C,D) -o
...}. When we first pick a fact for \texttt{a(A, B)} from the hash
table, we mark it as being used in order to ensure that, when we
retrieve facts for \texttt{a(C, D)}, the first one cannot be used
twice since that would violate linearity.

\begin{figure}[ht]
\centering
\includegraphics[width=0.4\textwidth]{figures/hash_table.pdf}
\caption{Hash table and doubly linked data structures for 
  a \texttt{a(int,int)} predicate}
\label{fig:hash_table}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Rule Engine}
\label{rule_engine}

The rule engine decides which rules may need to be executed while
taking into account rule priorities. Figure~\ref{fig:rule_engine}
shows the rule engine data structures in more detail. There are 4 main
data structures for scheduling rule execution: \texttt{Rule Queue} is
the bitmap representing the rules scheduled to run; \texttt{Active
Bitmap} contains the rules that have enough facts to be
fired; \texttt{Dropped Bitmap} contains the rules that must be dropped
from \texttt{Rule Queue} and \texttt{Active Bitmap} if the rule being
executed succeeds; and \texttt{Predicates Count} counts the number of
facts per predicate. To understand how our engine works, consider the
following set of rules and facts:

{\footnotesize
\begin{Verbatim}
// rules
a, e(1) -o b.
a -o c.
b -o d.
e(0) -o f.
c -o e(1).

// facts
a.
e(0).
\end{Verbatim}
}

\begin{figure*}[t]
\centering
\includegraphics[width=11cm]{figures/rule_queue.pdf}
\caption{Rule engine data structures (a) before and (b) after applying 
  the rule \texttt{a -o c.}}
\label{fig:rule_engine}
\end{figure*}

Since we have facts for predicates \texttt{a/0} and \texttt{e/1},
the \texttt{Active Bitmap} starts with rules 1, 2 and 4 marked as
having enough facts to be fired. The \texttt{Rule Queue} bitmap also
starts with the same three rules. In order to pick rules for
execution, we take the rule corresponding to the least significant bit
from the \texttt{Rule Queue} bitmap, initially the first
rule \texttt{a, e(1) -o b}. However, since we don't have
fact \texttt{e(1)}, this rule fails and we execute the second
rule \texttt{a -o c}. Figure~\ref{fig:rule_engine}(a) shows the rule
engine data structures at that point.

Because the derivation for the second rule succeeds, we will consume
fact \texttt{a} and derive fact \texttt{c}. We thus update
\texttt{Predicates Count} accordingly, mark the first 
and second rules in \texttt{Dropped Bitmap} since such rules are no
longer applicable (\texttt{a} was consumed) and mark the fifth rule
in \texttt{Active Bitmap} since \texttt{c} was derived. Finally, to
update the \texttt{Rule Queue}, we remove the bits marked
in \texttt{Dropped Bitmap} and add the active rules marked
in \texttt{Active Bitmap} that use the newly derived predicates, rule
5 in this case. In the continuation, the engine will schedule the
fourth and fifth rules to run. Figure~\ref{fig:rule_engine}(b) shows
the rule engine data structures at that point.

Note that every node in the program has the same set of data
structures presented in Fig.~\ref{fig:rule_engine}. We use 32 bits
integers to implement bitmaps and an array of 16 bits integers to
count facts, resulting in $32 + 2P$ bytes per node, where $P$ is the
number of predicates.

We do a small optimization to reduce the number of derivations of
persistent facts and, for that, we divide the program rules into two
sets: \emph{persistent rules} and \emph{non persistent rules}.
Persistent rules are rules where only persistent facts are
involved. We compile such rules incrementally, i.e., we attempt to
fire all rules where a persistent fact is used. This is called
the \emph{pipelined semi-naive} evaluation and it originated in the P2
system~\cite{Loo-condie-garofalakis-p2}. This evaluation method avoids
excessing re-derivations of the same fact. The order of derivation
does not matter for those rules, since only persistent facts are used.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Rule Execution}

@@

Consider the example in Fig.~\ref{fig:byte_code} that shows the LM
byte-code for the rule \texttt{!a(X,Y), b(X,Z), c(X,Y) -o
d(Y)}. Consider also a database with the
facts \texttt{!a(1,2)}, \texttt{!a(2,3)}, \texttt{b(1,3)},
\texttt{b(5,3)}, \texttt{c(1,2)}, \texttt{c(1,3)}, \texttt{c(5,3)}. Rule
execution for this rule and facts proceeds in a series of recursive
loops, as follows. The first loop retrieves an iterator for the
persistent facts of \texttt{!a/2} and moves the first valid
fact, \texttt{!a(1,2)}, to register 0. The inner loop retrieves linear
facts that match \texttt{b(1,Z)} (from the
\emph{join constraint}) and moves \texttt{b(1,3)} to register 1. The 
final loop moves \texttt{c(1,2)} to register 2 and the body of the
rule is successfully matched. Next, we derive \texttt{d(2)},
where \texttt{2} comes from register 0.

\begin{figure}[ht]
{\footnotesize
\begin{Verbatim}
PERSISTENT ITERATE a MATCHING TO reg 0
   LINEAR ITERATE b MATCHING TO reg 1
      (match).0=0.0             // match argument X
      LINEAR ITERATE c MATCHING TO reg 2
         (match).0=0.0          // match argument X
         (match).1=0.1          // match argument Y
         ALLOC d TO reg 3
         MVFIELDFIELD 0.1 TO 3.0  // get argument Y
         ADDLINEAR reg 3             // derive d(Y)
         REMOVE reg 2
         REMOVE reg 1
         TRY NEXT
      NEXT
   NEXT
RETURN
\end{Verbatim}
}
\caption{LM byte-code for rule \texttt{!a(X,Y), b(X,Z), c(X,Y) -o d(Y)}}
\label{fig:byte_code}
\end{figure}

In case of failure, we jump to the previous loop in order to try the
next candidate fact. In case of rule success, the head is derived and
we should backtrack to the inner most \emph{valid loop}, i.e., the
older loop that uses linear facts or, if there are no linear facts
involved, to the previous loop. We need to jump to a valid loop
because we may have loops with linear facts that are now invalid. In
our example, we would jump to the loop of \texttt{b(X,Z)} and
not \texttt{c(X,Y)}, since \texttt{b(1,3)} was consumed.

As an optimization, the compiler re-orders the fact expressions used
in the body in order to make execution more efficient. For example, it
forces the join constraints in rules to appear first so that matching
will fail sooner rather than later. It also does the same for
constraints. Note also that for every loop, the compiler adds
the \emph{match object}, which contains information about which
arguments need to match, so that runtime matching be efficient.

Our compiler also detects cases where we re-derive a linear fact with
new arguments. For example, as shown in Fig.~\ref{code:update}, the LM
byte-code for rule \texttt{a(N) -o a(N+1)} will compile to code that
reuses/updates the old \texttt{a(N)} fact when deriving the
new \texttt{a(N+1)} fact.

\begin{figure}[ht]
{\footnotesize
\begin{Verbatim}
LINEAR ITERATE a MATCHING TO reg 0
  MVFIELDREG 0.0 TO reg 1       // initial argument
  MVINTREG INT 1 TO reg 2
  reg 1 INT PLUS reg 2 TO reg 3
  MVREGFIELD reg 3 TO 0.0  // reuse/update argument
  UPDATE reg 0
  TRY NEXT
RETURN
\end{Verbatim}
}
\caption{\small{LM byte-code for rule \texttt{a(N) -o a(N+1)}}}
\label{code:update}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Indexing}
\label{indexing}

@@

{\tiny
\[
Entropy(A, F) = - \sum_{v \in values(F, A)} \frac{count(F, A = v)}{total(F)} \log_2 \frac{count(F, A = v)}{total(F)}
\]
}

@@

For one of the arguments to score, $Entropy(A, F)$ multiplied by the
number of times it has been used for lookup must be larger than the
other argument. The argument with the best score is selected and then
a global variable called \texttt{indexing\_epoch} is updated.  In
order to convert the node's linked lists into hash tables, each node
also has a local variable called \texttt{indexing\_epoch} that is
compared to the global variable in order to rebuild the node database
according to the new indexing information.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

LM also supports recursive types such as lists and pairs. These
complex data structures are stored in the heap of the VM and are
managed through reference counting. For instance, each list is
a \emph{cons cell} with 3 fields: \texttt{tail}, the pointer to the
next element of the list; \texttt{head}, the element stored by this
element of the list; and \texttt{refs} that counts the number of
pointers to this list element in the VM. The list is deleted from the
heap whenever \texttt{refs} is decremented to 0.
